---
title: "Chimeric Pho4 data analyzer v3"
author: "Bin He"
date: '2023-11-26'
output:
  html_notebook:
    code_folding: hide
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r echo=FALSE, message=FALSE}
suppressPackageStartupMessages(require(shinyWidgets))
require(bslib)
require(bsicons)
suppressPackageStartupMessages(require(DT))
suppressPackageStartupMessages(require(plotly))
suppressPackageStartupMessages(require(tidyverse))
suppressPackageStartupMessages(require(cowplot))
suppressPackageStartupMessages(require(ggtext))
require(grid)
require(RColorBrewer)
```

Flow cytometry data is generated by Lindsey Snyder between 2023.03-05. For QC plots, see [here](https://rpubs.com/emptyhb/d231027-LFS-PHO5-qc-and-pool).

```{r data}
# Import the background subtracted data. Filter.
dat0 <- read_tsv("20231023-PHO5-bg-subtracted-data.tsv", col_types = "ccccdddddc")

dat <- filter(dat0, host != "PHO84", flag == "pass", date != "02/10") %>% 
  # based on previous QC, the following sample (both replicates) have high
  # variance - one biological replicate is highly expressed, while the other 
  # two have mNeon, but barely any RFP expression.
  mutate(flag = ifelse(plasmid == "233" & host == "pho2∆", "high.var", flag))

# load summarized chimera data
ximera <- read_tsv("20231125-PHO5pr-chimera-summarized.tsv", col_types = "cccfddddd")

# get the symbols of all chimeras and sort them for selection
ximeras <- sort(ximera$symbol)
```

```{r parameters}
# Set up common parameters for thresholding and plotting
# reference Pho4 plasmid ids
refs <- c("188", "194")
# colors
date.colors = c(brewer.pal(name="Dark2", n = 8), brewer.pal(name="Paired", n = 8))
point.colors = c("PHO2" = "gray10", "pho2" = "gray30")
host.colors = c("PHO2" = "gray50", "pho2" = "orange2")
```


```{r functions}
# Define functions
# 1. Select data
my_data_select <- function(pattern = NULL, Set = NULL){
  # function to select a subset of the chimeras given a set of rules
  # validate the input
  
  # change region 4 into a consistent format
  tmp <- ximera %>% 
    mutate(
      symbol = as.character(symbol),
      Symbol = ifelse(
        nchar(symbol) == 5,
        paste0(str_sub(symbol, 1, 3), 
               str_sub(symbol, 4, 4), 
               str_sub(symbol, 4, 4), 
               str_sub(symbol, 5, 5)),
        symbol
      )) %>% 
    select(plasmid, Symbol, set)
  # starting set
  if(length(Set) == 0)
    xim <- filter(tmp, !plasmid %in% refs)
  else
    xim <- filter(tmp, set %in% Set, !plasmid %in% refs)
  # compare to the pattern
  try(if(nchar(pattern) != 6) stop("Pattern must be a string with 6 characters"))
  symbols = xim$Symbol # extrac the symbols for testing
  include = nchar(symbols) > 0 # initialize the inclusion vector
  for(i in 1:6){
    p = substr(pattern, i, i)
    if(p != "X" & p != "x"){ # ignore X and x
      test = toupper(str_sub(symbols, i, i)) == toupper(p)
      include = include & test
    }
  }
  select <- cbind(xim, include)
  return(select$plasmid[select$include])
}

my_data_select_m <- function(patterns = NULL, Set = NULL){
  # combine multiple selections
  all_selected = c()
  try(if(length(patterns) == 0) stop("No patterns provided"))
  for(i in patterns)
    all_selected = c(all_selected, my_data_select(pattern = i))
  return(unique(all_selected))
}

my_data_add <- function(input){
  # given a comma separated set of symbols, return the corresponding plasmid IDs
  symbols <- unlist(str_split(input, pattern = "\\s*,\\s*"))
  ximera$plasmid[ximera$symbol %in% symbols]
}

my_selection_combine <- function(set1, set2){
  # given two vectors, combine them and return the unique set of plasmid IDs
  unique(c(set1, set2))
}

my_calc_region_effect <- function(region, variable){
  # this function takes the name of a variable of interest
  # x specifies the foreground region, which will be examined for its effect on
  # the variable of interest.
  # it then transforms the ximera data frame to preserve only the variable of
  # interest, pivots it wider after grouping by the background composition.
  
  # prepare the data by mutating the symbol column into fg and bg
  valid.var <- c("A_PHO2", "A_pho2", "s_PHO2", "spho2", "boost")
  if(!variable %in% valid.var)
    stop(paste0("Please specify one of the valid variable names:", 
                paste(valid.var, collapse = ", ")))
  tmp <- ximera %>% 
    filter(set == "M") %>% 
    select(plasmid, symbol, var = {{ variable }}) %>% 
    mutate(fg = str_sub(symbol, region, region) %>% toupper(),
           bg = symbol %>% toupper())
  # replace the foreground region with X for grouping
  str_sub(tmp$bg, region, region) <- "X"
  # reorganize the tibble for easier handling, optional
  tmp <- relocate(tmp, fg, bg, .before = symbol) %>% select(-symbol)
  # pivot the data into a wide format such that for each background, there
  # are two values for the variable of interest, one from the chimera with 
  # CgPho4's version in the foreground and another with ScPho4's version
  tmp <- tmp %>% 
    select(plasmid, fg, bg, var) %>% 
    pivot_wider(id_cols = bg, names_from = "fg", 
                values_from = c(plasmid, var)) %>% 
    unite(plasmid, starts_with("plasmid")) %>%
    mutate(label = paste(bg, plasmid, sep = "\n"))
  return(tmp)
}
# 2. Data prep and transform
my_data_prep <- function(selection){
  # given a selection of chimera ID (plasmid), prepare a data frame for plotting
  # subset data
  tmp <- ximera %>% 
    filter(plasmid %in% c(refs, selection)) %>% 
    select(plasmid, symbol, group) %>% 
    inner_join(dat, by = "plasmid") %>% 
  return(tmp)
}

my_comp_region_effect <- function(region){
  # this function uses my_calc_region_effect to get the value for the variable of interest
  # with either Cg or Sc version in the focal region, separately for each background composition
  # it does so for two variables, A_PHO2 and A_pho2, then calculate dA_PHO2, dA_pho2, and
  # combine them
  PHO2 = my_calc_region_effect(region, "A_PHO2") %>% 
    mutate(dA_PHO2 = var_C - var_S,
           # mean A_PHO2
           M_PHO2 = (var_S + var_C)/2,
           NF = ifelse(M_PHO2 <=3.5, TRUE, FALSE)) %>% 
    select(-var_S, -var_C)
  
  pho2 = my_calc_region_effect(region, "A_pho2") %>% 
    mutate(dA_pho2 = var_C - var_S, 
           M_pho2 = (var_S + var_C)/2) %>% 
    select(-var_S, -var_C)
  
  dat <- full_join(PHO2, pho2, by = c("bg", "plasmid", "label")) %>% 
    select(bg, plasmid, dA_PHO2, dA_pho2, M_PHO2, M_pho2, NF)
  
  return(dat)
}

my_upper_triangular_mat <- function(alt = "C", var = "A_PHO2"){
  # given the alternative allele (C/S) and a variable of interest, e.g., A_PHO2,
  # output an upper triangular matrix containing the values from the variable 
  # of interest, with the row and col numbers based on the first and second
  # positions containing the alternative allele. If all positions contain the 
  # reference allele, the value is subtracted from all values in the matrix
  # when just one position is the alternative allele, the value in the diagonal
  # is set. when there are more than 2 regions containing the alternative allele
  # skip.
  out_mat <- matrix(NA, nrow = 5, ncol = 5)
  ref_val <- NA
  dat <- filter(ximera, set == "M") %>% 
    mutate(S = as.character(symbol) %>% toupper())
  for(i in seq(1, nrow(dat))){
    symbol = dat[i, "S"]
    # determine which positions contain the alternative allele
    p = str_locate_all(symbol, alt)[[1]][,"start"]
    l = length(p)   # how many positions contain the alt allele
    v = dat[[var]][i] # retrieve the value of the variable
    if(l == 0)
      ref_val = v
    else if(l == 1)
      out_mat[p, p] = v
    else if(l == 2)
      out_mat[p[1], p[2]] = v
  }
  out_mat = out_mat - ref_val
  return(out_mat)
}

my_combined_triangular_mat <- function(alt = "C"){
  # given the alternative allele (C/S), output a matrix containing the values
  # for both with and without Pho2, arranged in two complementary triagular
  # matrices, with the row and col numbers based on the first and second
  # positions containing the alternative allele. If all positions contain the 
  # reference allele, the value is subtracted from all values in the matrix
  # when just one position is the alternative allele, the value in the diagonal
  # is set. when there are more than 2 regions containing the alternative allele
  # skip.
  out_mat <- matrix(NA, nrow = 6, ncol = 6)
  upper <- cbind(NA, my_upper_triangular_mat(alt, var = "A_PHO2")) %>% 
    rbind(., NA)
  lower <- rbind(NA, t(my_upper_triangular_mat(alt, var = "A_pho2"))) %>% 
    cbind(., NA)
  out_mat = ifelse(is.na(upper), lower, upper)
  return(out_mat)
}

# 3. Plotting
themes <- list(
  labs(caption = "ref = reference; n.f. = non functional"),
  theme_cowplot(font_size = 20),
  panel_border(color = "gray30"),
  background_grid(major = "y", minor = "none"),
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = rel(0.9)),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        strip.placement = "outside",
        strip.background = element_blank(),
        strip.text.y.left = element_markdown())
)

my_plot_ratio <- function(selection){
  # plot R/G ratios and the individual components
  # this is the "raw data plot" for examining the underlying evidence
  # for higher level plots like the relative activity below
  tmp <- my_data_prep(selection)
  p <- tmp %>% 
    select(-c(FSC.H, nGFP, nRFP, flag)) %>% 
    mutate(`R/G` = YL2.H/BL1.H) %>% 
    pivot_longer(cols = c(BL1.H, YL2.H, `R/G`), 
                 names_to = "parameter", values_to = "value") %>% 
    mutate(parameter = factor(parameter, levels = c("R/G", "YL2.H", "BL1.H"),
                              labels = c("RFP/GFP", "PHO5pRFP", "Pho4-GFP")),
           host = fct_recode(host, pho2 = "pho2∆")) %>% 
    ggplot(aes(x = symbol, y = value, group = host)) + 
    stat_summary(aes(group = host), fun.data = "mean_cl_boot", geom = "errorbar",
                 position = position_dodge(0.6), width = 0.3) +
    geom_bar(aes(fill = host), alpha = 0.8, width = 0.6,
             stat = "summary", fun = "mean", position = position_dodge(0.6)) +
    geom_point(data = function(x) subset(x, group != "ref"),
               aes(group = host, color = host), size = 1, shape = 3, alpha = 0.9,
               position = position_jitterdodge(dodge.width = 0.7, jitter.width = 0.1)) +
    scale_color_manual(values = point.colors) +
    scale_fill_manual(values = host.colors) +
    facet_grid(parameter~group, scales = "free", space = "free_x", switch = "y") +
    xlab("Pho4 chimera") + themes + theme(legend.position = "top")
  return(p) 
}

my_plot_boost <- function(selection){
  # given a selection of chimera IDs, plot their functionality w/PHO2
  # relative to ScPho4, and their boost
  # dat
  tmp <- filter(ximera, plasmid %in% c(refs, selection)) %>% 
    mutate(perc_pho2 = A_pho2/A_PHO2, 
           symbol = paste(symbol, set, sep = "-")) %>% 
    pivot_longer(cols = c(s_PHO2, boost, s_pho2), 
                 names_to = "parameter", values_to = "ratio")
  # labeller
  par.explain <- c(
    s_PHO2 = "Rel. A<sub>PHO2</sub>",
    s_pho2 = "Rel. A<sub>pho2</sub>",
    boost = "Boost",
    perc_pho2 = "%A<sub>pho2</sub>"
  )
  p <- ggplot(tmp, aes(x = symbol, y = ratio)) +
    geom_col(width = 0.5, color = "black", fill = "gray80") +
    geom_hline(yintercept = 1, linetype = 2, color = "gray30") +
    facet_grid(parameter~group, scales = "free", space = "free_x", switch = "y",
              labeller = labeller(parameter = par.explain)) +
    #theme_bw(base_size = 18) +
    #background_grid(minor = "none") +
    themes
  return(p)
}

my_scatter_plot <- function(pattern, Set = NULL){
  # plot all chimeras by their A_PHO2 and A_pho2∆, highlighting a subset of
  # the chimeras by their composition.
  selection = my_data_select(pattern = pattern, Set = Set)
  scatter.colors = c("ScPho4" = "forestgreen", "CgPho4" = "blue3", "cyan2", "other" = "gray20")
  names(scatter.colors)[3] = pattern
  p <- ximera %>% 
    mutate(A_PHO2 = signif(A_PHO2, digits = 2),
           A_pho2 = signif(A_pho2, digits = 2),
           group = case_when(
             symbol == "CCCCC" ~ "CgPho4",
             symbol == "SSSSS" ~ "ScPho4",
             plasmid %in% selection ~ pattern,
             .default = "other"
           ),
           group = fct_relevel(group, names(scatter.colors))) %>% 
    ggplot(aes(x = A_PHO2, y = A_pho2, label = symbol)) + 
    geom_point(aes(color = group), size = 2.5) + 
    scale_color_manual("Pho4 type", values = scatter.colors) +
    geom_abline(slope = 1) +
    theme_gray(base_size = 14)
  return(p)
}

my_plot_region_effect_twovar_line <- function(region, highlight = "none"){
  # this function uses my_comp_region_effect to generate the data
  # and plot the difference in A_PHO2 and A_pho2 between the CgPho4 vs ScPho4
  # in the focal region. an optional highlight option allows one other region
  # to be colored based on the species of origin
  dat <- my_comp_region_effect(region) %>% 
    pivot_longer(cols = c(dA_PHO2, dA_pho2), 
                 names_to = "host", values_to = "diff") %>% 
    mutate(host = fct_recode(host, `PHO2` = "dA_PHO2", `pho2∆` = "dA_pho2"),
           host = fct_relevel(host, "PHO2"))
  legend.title = ""
  if(highlight != "none" & highlight != region){
    hl = as.numeric(highlight)
    dat <- mutate(dat, grp = str_sub(bg, hl, hl) %>% toupper(),
                  grp = fct_recode(grp, CgPho4 = "C", ScPho4 = "S"))
    legend.title = paste("Region", highlight, sep = " ")
  }else{
    dat <- mutate(dat, grp = ifelse(NF, "no", "yes"))
    legend.title = "Functional"
  }
  p <- dat %>% 
    ggplot(aes(x = host, y = diff, label = bg)) +
    geom_point(aes(color = grp), size = 2, alpha = 0.8,
               position = position_jitter(0.05)) + 
    geom_line(aes(group = bg, color = grp), linewidth = 0.2, alpha = 0.8) +
    scale_color_manual(legend.title, values = c("orange", "gray10")) +
    ylab(paste("Region", region, "swap", "(Cg-Sc)", sep = " ")) +
    theme_bw(base_size = 18) + 
    theme(
      axis.title.x = element_blank(),
      legend.text = element_text(size = rel(0.8)),
      legend.title = element_text(size = rel(0.9), face = 3)
    )
  return(p)
}

my_plot_combined_triangle_heatmap <- function(alt){
  # this function takes the output of the function my_combined_triangular_mat()
  # using pheatmap function, then rotates it using grid graphics
  # thanks to https://bookdown.org/rdpeng/RProgDA/the-grid-package.html#grid-graphics-coordinate-systems
  # adding title based on https://davetang.github.io/muse/pheatmap.html
  
  # construct title of plot
  ref = ifelse(alt == "C", "ScPho4", "CgPho4")
  my_title <- paste("Epistasis between regions on", ref, "background")
  test <- my_combined_triangular_mat(alt = alt)
  paletteLength = 50
  myColors <- colorRampPalette(c("steelblue4", "gray90", "red"))(paletteLength)
  rng <- max(abs(test), na.rm = TRUE)
  myBreaks <- c(seq(-rng, 0, length.out=ceiling(paletteLength/2) + 1), 
                seq(rng/paletteLength, rng,
                    length.out=floor(paletteLength/2)))
  p <- pheatmap::pheatmap(test, color = myColors, breaks = myBreaks,
                          cellwidth = 40, cellheight = 40,
                          border_color = NA, na_col = NA, silent = TRUE,
                          cluster_cols = FALSE, cluster_rows = FALSE)
  vp <- viewport(x = 0.5, y = 0.45,
                 width = unit(3, "in"), height = unit(3, "in"), angle = 45) 
  grid.newpage()
  pushViewport(vp)
  grid.draw(p$gtable)
  popViewport()
  grid.text(label = my_title, x = 0.5, y = 0.95, 
            gp = gpar(fontsize = 16, fontface = "bold"))
  grid.text(label = "With Pho2", x = 0.2, y = 0.75, just = "center",
            gp = gpar(fontsize = 14, fontface = "bold"))
  grid.text(label = "Without pho2", x = 0.2, y = 0.15, just = "center", 
            gp = gpar(fontsize = 14, fontface = "bold"))
  return(p)
}
```

```{r}
help = '
Enter patterns as "XXXXXX". Position corresponds to one of the 6 regions to be specified (4,5 = region 4.1 and 4.2, 6 = region 5).
Three characters are allowed - "X" = either parental Pho4 orthologs; "S" = ScPho4 and "C" = CgPho4.
Example: "XXXCCX" = all chimeras with region 4 completely from CgPho4
'

page_fillable(
  card(
    card_header(
      "Selecting Pho4 chimeras for plotting",
      popover(
        bsicons::bs_icon("question-circle"),
        help,
        placement = "right"
      ),
      class = "d-flex align-items-center gap-1"
    ),
    card_body(
      layout_column_wrap(
        width = 1/2,
        textInput(inputId = "pattern", label = "Please enter a pattern, e.g., 'XXCXXC'", value = "CXXXXC"),
        checkboxGroupInput(inputId = "set", label = "Please select the sets to include",
                           choiceNames = c("Main (M)", "P2ID_split (S)", "Other (B)"), 
                           choiceValues = c("M", "S", "B"),
                           selected = c("M", "S", "B"),
                           inline = TRUE),
        gap = "5%"
      )
    )
  )
)
# validate the input
renderText({
  validate(
    need(grepl("^[CSX]{6}$", input$pattern, ignore.case = TRUE),
         "Main pattern is invalid!")
  )
})
```

### Results {.tabset}
#### Relative activity
```{r}
# reactive function to produce the plot
plot_boost <- reactive({
    selection = my_data_select(pattern = input$pattern, Set = input$set)
    p <- my_plot_boost(selection)
    return(p)
})

fluidPage(
  # App title ----
  
  # Output: Tabset w/ plot, summary, and table ----
  downloadHandler(
    filename = paste0(gsub('-', '', Sys.Date()),
                      "Pho4-chimeras-Pho2dep-", 
                      sample(1:100, 1), ".pdf"),
    content = function(file){
      #p <- my_plot_boost(my_data_select(input$pattern))
      ggsave(file, plot_boost(), device = "pdf", width = 10, height = 7)
    }
  ),
  #renderPrint({input$set}),
  renderPlot({
    plot_boost()
  },
  height = 500)
)
```

<span style="color: red;align: center">**Note**</span> 

**n.f.** = "non functional", meaning activity with Pho2 is less than 20% of ScPho4 with Pho2.

**boost** = A<sub>PHO2</sub>/A_<sub>pho2∆</sub>; **Rel. A<sub>PHO2</sub>** = A<sub>PHO2, chimera</sub>/A<sub>PHO2, ScPho4</sub>; **Rel. A<sub>pho2</sub>** = A<sub>pho2, chimera</sub>/A<sub>pho2</sub>;

#### Components
```{r}
plot_ratio <- reactive({
    selection = my_data_select(input$pattern, input$set)
    p <- my_plot_ratio(selection)
    return(p)
})


fluidPage(
  # App title ----
  
  # Output: Tabset w/ plot, summary, and table ----
  downloadHandler(
    filename = paste0(gsub('-', '', Sys.Date()),
                      "-Pho4-chimeras-QC-",
                      sample(1:100, 1), ".pdf"),
    content = function(file){
      ggsave(file, plot_ratio(), device = "pdf", width = 10, height = 7)
    }
  ),
  renderPlot({
    plot_ratio()
    #print(p)
    #set_null_device("png") 
    #plot_grid(p1, p2, p3, ncol = 1, rel_heights = c(6,4,3))
  },
  height = 500)
)
```

<span style="color: red;align: center">**Note**</span>  **n.f.** = "non functional", meaning activity with Pho2 is less than 20% of ScPho4 with Pho2.

#### Scatter plot
```{r}
renderPlotly({
  p4 <- my_scatter_plot(input$pattern, input$set)
  ggplotly(p4, tooltip = c("label", "x", "y")) %>% 
    config(toImageButtonOptions = 
             list(format = "png",
                  filename = paste0(gsub('-', '', Sys.Date()),
                                    "-Pho4-chimeras-scatter-",
                                    sample(1:100, 1)),
                  height = 500,
                  width = 700,
                  scale = 2
             ))
})
```

#### Region swap effect
```{r}
help1 = '
Select one region as the focal region. The difference between paired chimeras differing only in that region but identical in the rest of the protein are plotted. The corresponding values with or without Pho2 are connected by a line.
'
help2 = '
Select another region to color the dots by the Pho4 origin at that region
'

fluidPage(
  sidebarLayout(
    sidebarPanel(
  prettyRadioButtons( # or radioGroupButtons
    inputId = "focalregion",
    label = tooltip(
      trigger = list(
        "Focal region",
        bs_icon("info-circle")
      ),
      help1
    ),
    choices = c("1", "2", "3", "4", "5"),
    bigger = TRUE
  ), 
  sliderTextInput(
    inputId = "highlight",
    label = tooltip(
      trigger = list(
        "Group by",
        bs_icon("info-circle")
      ),
      help2
    ),
    choices = c("none", "1", "2", "3", "4", "5"),
    selected = "none"
  ),
  width = 3),
  mainPanel(
  renderPlotly({
    my_plot_region_effect_twovar_line(input$focalregion, input$highlight) %>% 
      ggplotly() %>% 
      config(toImageButtonOptions = 
               list(format = "png",
                    filename = paste0(gsub('-', '', Sys.Date()),
                                      "-Pho4-chimeras-region-",
                                      sample(1:100, 1)),
                    height = 400,
                    width = 500,
                    scale = 2.5
               ))
  }), width = 9)
  ),
)
```
#### Triangle heatmap

```{r}
fluidPage(
  fluidRow(
    column(
      6,
      renderPlot({
        p1 <- my_plot_combined_triangle_heatmap("C")
      }),
      downloadHandler(
        filename = paste0(gsub('-', '', Sys.Date()),
                          "-triangle-heatmap-ScPho4-ref-",
                          sample(1:100, 1), ".pdf"),
        content = function(file){
          pdf(file, width = 7, height = 6)
          my_plot_combined_triangle_heatmap("C")
          dev.off()
        }
      )
      
    ),
    column(
      6,
      renderPlot({
        p2 <- my_plot_combined_triangle_heatmap("S")
      }), 
      
      downloadHandler(
        filename = paste0(gsub('-', '', Sys.Date()),
                          "-triangle-heatmap-CgPho4-ref-",
                          sample(1:100, 1), ".pdf"),
        content = function(file){
          pdf(file, width = 7, height = 6)
          my_plot_combined_triangle_heatmap("S")
          dev.off()
        }
      )
    )
  )
)
```
> In the heatmaps above, the cells at the base of each triangle shows the effect of one region swaps (linearly arranged from 1-5) on the indicated Pho4 background. The rest of the triangle shows the two region swap effects. The values used to plot the color gradients are the difference in the respective chimera and the reference Pho4. For example, the tip of the top triangle has a value of A<sub>PHO2, CSSSC</sub> - A<sub>PHO2, SSSSS</sub>

#### Chimera lookup
```{r}
ximera %>% 
  mutate(across(A_PHO2:boost, \(x) signif(x, digits = 2))) %>% 
  renderDT(rownames = FALSE)
```
