---
title: merge median intensity data for gcPBM
author: Bin He
date: 2024-04-18
output: html_notebook
---

# Goal:
Yuning gave us a modified perl script, which actually does the detrending and calculates the median intensity for each probe.

```{r, message=FALSE}
require(tidyverse)
require(cowplot)
require(ggpubr)
```

# Date wrangling
Load files into memory
```{r, message=FALSE}
rawSc = read_tsv("../input/normalized-intensities/20240418-ScHigh_alldata_ctrl_M_removed.txt.gz")
rawCg = read_tsv("../input/normalized-intensities/20240418-CgHigh_alldata_ctrl_M_removed.txt.gz")
```

reformat the datasets for ScPho4
```{r}
name_pattern = c(group = "^[^_]*_[^_]*", "_",  name = ".*", rep = "_o\\d_r\\d$")
name_pattern_neg = c(group = "Cg_neg", "_", name = "\\d+", "_", rep = "\\d$")
datSc <- rawSc %>% 
  filter(!grepl("Cg_neg", Name), !is.na(Name)) %>% 
  # split the name column to create a group and a probe name
  separate_wider_regex(Name, patterns = name_pattern, too_few = "debug") %>% 
  # I manually inspected, and found all the Name_ok == FALSE ones are 
  # Agilent control probes, like DarkCorner. remove those are fine
  filter(Name_ok) %>% 
  mutate(
    consensus = grepl("CACGTG", Sequence),
    group = ifelse(group == "sc_Probes",
                   ifelse(consensus, "Sc_consensus", "Sc_nonconsensus"),
                   group)) %>% 
  select(ID, group, name, rep,
         sequence = Sequence,
         starts_with("Alexa"))

datSc <- rbind(
  datSc,
  filter(rawSc, grepl("Cg_neg", Name)) %>% 
  # split the name column to create a group and a probe name
  separate_wider_regex(Name, patterns = name_pattern_neg) %>% 
  select(ID, group, name, rep,
         sequence = Sequence,
         starts_with("Alexa"))
)

medSc <- datSc %>% 
  group_by(group, name) %>% 
  summarize(
    sequence = first(sequence),
    my_median = median(Alexa488Adjusted)
  )
```

reformat the datasets for CgPho4
```{r}
datCg <- rawCg %>% 
  filter(!grepl("Cg_neg", Name), !is.na(Name)) %>% 
  # split the name column to create a group and a probe name
  separate_wider_regex(Name, patterns = name_pattern, too_few = "debug") %>% 
  filter(Name_ok) %>% 
  mutate(
    consensus = grepl("CACGTG", Sequence),
    group = ifelse(group == "sc_Probes",
                   ifelse(consensus, "Sc_consensus", "Sc_nonconsensus"),
                   group)) %>% 
  select(ID, group, name, rep,
         sequence = Sequence,
         starts_with("Alexa"))

datCg <- rbind(
  datCg,
  filter(rawCg, grepl("Cg_neg", Name)) %>% 
  # split the name column to create a group and a probe name
  separate_wider_regex(Name, patterns = name_pattern_neg) %>% 
  select(ID, group, name, rep,
         sequence = Sequence,
         starts_with("Alexa"))
)

medCg <- datCg %>% 
  group_by(group, name) %>% 
  summarize(
    sequence = first(sequence),
    my_median = median(Alexa488Adjusted)
  )
```

join the two datasets
```{r}
medJoin <- inner_join(medCg, medSc, by = c("group", "name", "sequence"), 
                      suffix = c("_Cg", "_Sc"))
medJoin$group = factor(medJoin$group,
                       levels = c("BLI_UASp2", "MitomiProbe_mut", "Cg_consensus",
                                  "Cg_nonconsensus", "Sc_consensus",
                                  "Sc_nonconsensus", "sc_NegCtrl", "Cg_neg"),
                       labels = c("BLI_UASp2", "MitomiProbe_mut", "Cg_consensus",
                                  "Cg_nonconsensus", "Sc_consensus", 
                                  "Sc_nonconsensus", "Sc_negative", "Cg_negative")
                       )
table(medJoin$group)
```

# Plot
Scatter plot for both consensus and nonconsensus sequences
```{r}
grp.color = c("Sc_consensus" = "forestgreen", "Sc_nonconsensus" = "palegreen",
              "Cg_consensus" = "blue2", "Cg_nonconsensus" = "skyblue",
              "Sc_negative" = "gray30", "Cg_negative" = "gray")
#grp.color = c("consensus" = "palegreen", "nonconsensus" = "forestgreen",
#              "negative" = "gray30")
p <- medJoin %>% 
  filter(grepl("Sc|Cg", group)) %>%
  #separate(group, into = c("species","group"), sep = "_") %>% 
  mutate(lnSc = log(my_median_Sc), lnCg = log(my_median_Cg),
         sps = str_sub(group, 1, 2)) %>% 
  ggplot(aes(x = lnSc, y = lnCg)) +
  geom_point(aes(color = group, alpha = sps, size = sps), shape = 19) +
  scale_alpha_manual(values = c("Sc" = 0.5, "Cg" = 1), guide = "none") +
  scale_size_manual(values = c("Sc" = 1, "Cg" = 1.3), guide = "none") +
  #geom_abline(slope = 1, intercept = 0, color = "red", linetype = 3) +
  scale_color_manual("Probe group:", values = grp.color) +
  xlab("ScPho4 log signal intensity") +
  ylab("CgPho4 log signal intensity") +
  theme_cowplot() + panel_border(color = "gray30", size = 1.5) +
  theme(
    axis.line = element_blank(),
    axis.text = element_text(size = rel(1.1)),
    legend.position = c(0.05, 0.75),
    legend.title = element_text(size = rel(0.7)),
    legend.text = element_text(size = rel(0.6)),
    legend.background = element_rect(fill = alpha("white", 0.5))
  )
p + stat_cor(method = "spearman", label.x.npc = 0.6, label.y.npc = "bottom")
ggsave("../output/20240418-gcPBM-genomic-probes-and-nc-only.png", width = 4.2, height = 4)
```

Only show sequences with the E-box consensus
```{r}
grp.color.sub = c("Sc_consensus" = "forestgreen", "Cg_consensus" = "blue2")
              #"Sc_negative" = "gray30", "Cg_negative" = "gray")
#grp.color = c("consensus" = "palegreen", "nonconsensus" = "forestgreen",
#              "negative" = "gray30")
tmp <- medJoin %>% 
  filter(
    grepl("Sc|Cg", group),
    grepl("_consensus", group),
  ) %>%
  #separate(group, into = c("species","group"), sep = "_") %>% 
  mutate(lnSc = log(my_median_Sc), lnCg = log(my_median_Cg),
         sps = str_sub(group, 1, 2))

corr <- with(tmp, cor(lnSc, lnCg, method = "spearman"))
p <- tmp %>% 
  ggplot(aes(x = lnSc, y = lnCg)) +
  geom_point(aes(color = group), alpha = 0.9, shape = 16, size = 0.7) +
  #geom_abline(slope = 1, intercept = 0, color = "steelblue") +
  #scale_alpha_manual(values = c("Sc" = 0.5, "Cg" = 1), guide = "none") +
  #scale_size_manual(values = c("Sc" = 1, "Cg" = 1.3), guide = "none") +
  #geom_abline(slope = 1, intercept = 0, color = "red", linetype = 3) +
  scale_color_manual("Genomic context:", values = grp.color.sub,
                     labels = c("S. cerevisiae", "C. glabrata")) +
  xlab("ScPho4 log signal intensity") +
  ylab("CgPho4 log signal intensity") +
  theme_cowplot() + panel_border(color = "gray30", size = 1.5) +
  theme(
    axis.line = element_blank(),
    axis.text = element_text(size = rel(1.1)),
    legend.position = "top",
    legend.title = element_text(size = rel(0.8)),
    legend.text = element_text(size = rel(0.6), face = 3)
  )
p + stat_cor(method = "spearman")
ggsave("../output/20241007-gcPBM-genomic-consensus-only.png", width = 4, height = 4)
```

_**Discussion**_

The absolute values of the log intensity are not directly comparable, as that can be influenced by the effective concentrations of the two proteins hybridized on the chips. Instead, pay attention to the Spearman's correlation coefficient. A high level of correlation indicates very similar specificities.

Only show sequences with  non-consensus motifs
```{r}
grp.color.sub = c("Sc_nonconsensus" = "forestgreen", "Cg_nonconsensus" = "blue2")
              #"Sc_negative" = "gray30", "Cg_negative" = "gray")
#grp.color = c("consensus" = "palegreen", "nonconsensus" = "forestgreen",
#              "negative" = "gray30")
tmp <- medJoin %>% 
  filter(
    grepl("Sc|Cg", group),
    grepl("_nonconsensus", group),
  ) %>%
  #separate(group, into = c("species","group"), sep = "_") %>% 
  mutate(lnSc = log(my_median_Sc), lnCg = log(my_median_Cg),
         sps = str_sub(group, 1, 2))

corr <- with(tmp, cor(lnSc, lnCg, method = "spearman"))
p <- tmp %>% 
  ggplot(aes(x = lnSc, y = lnCg)) +
  geom_point(aes(color = group, alpha = sps), shape = 16, size = 0.7) +
  #geom_abline(slope = 1, intercept = 0, color = "steelblue") +
  scale_alpha_manual(values = c("Sc" = 0.5, "Cg" = 1), guide = "none") +
  #scale_size_manual(values = c("Sc" = 1, "Cg" = 1.3), guide = "none") +
  #geom_abline(slope = 1, intercept = 0, color = "red", linetype = 3) +
  scale_color_manual("Genomic context:", values = grp.color.sub,
                     labels = c("S. cerevisiae", "C. glabrata")) +
  xlab("ScPho4 log signal intensity") +
  ylab("CgPho4 log signal intensity") +
  theme_cowplot() + panel_border(color = "gray30", size = 1.5) +
  theme(
    axis.line = element_blank(),
    axis.text = element_text(size = rel(1.1)),
    legend.position = "top",
    legend.title = element_text(size = rel(0.8)),
    legend.text = element_text(size = rel(0.6), face = 3)
  )
p + stat_cor(method = "spearman")
ggsave("../output/20241008-gcPBM-genomic-nonconsensus-only.png", width = 4, height = 4)
```