---
title: "Plot PADDLE predictions"
date: 2024-10-29, updated `r Sys.Date()`
author: Bin He
format: html
editor: visual
---

```{r}
#| message: false
require(tidyverse)
require(cowplot)
```

## Goal

Plot PADDLE predicted activation potentials for multiple Pho4 orthologs

## Import data

```{r}
my_read_paddle <- function(name){
  # read the file
  tmp <- read_csv(name, col_types = cols())
  # extend the scores to the two ends
  tmp <- tmp[c(rep(1, 26), 1:nrow(tmp), rep(nrow(tmp), 26)),] %>% 
    mutate(Pos = 1:n(), rel_pos = Pos/n())
}

files = dir("../output/", pattern = "20240316*")
names(files) <- str_split(files, pattern = "-", simplify = TRUE)[,2]
dat <- map_dfr(files, \(f) my_read_paddle(paste0("../output/", f)), .id = "Pho4")
```

## Pho4 annotation and phylogeny
In this analysis, we included 14 Pho4 orthologs. It would be useful to have a common set of annotations for them in BED format, including the location of the bHLH domain and coordinates and lengths of the P2ID. We would also want to have the phylogeny for the species from which the Pho4 orthologs come.

```{r}
anno <- read_csv("../data/annot/20241029-Pho4-ortholog-annotation.csv", 
                 comment = "#") %>% 
  mutate(len = end - start + 1)
```
## Plotting
```{r}
act.colors <- c("gray80", "royalblue", "royalblue3")
names(act.colors) <- c("no act", "medium", "strong")
#select.sps <- c("ScerPho4", "SmikPho4", "CglaPho4", "CbraPho4", "LkluPho4", "CalbPho4")
p3 <- dat %>% 
  #filter(Pho4 %in% select.sps) %>% 
  mutate(Pho4 = factor(Pho4, levels = rev(select.sps)),
         paddle = cut(Z_score, breaks = c(-0.5, 4, 6, 12), 
                      labels = c("no act", "medium", "strong"))) %>% 
  ggplot(aes(x = Pos, y = Pho4, height = h)) +
  geom_tile(aes(fill = paddle), width = 2, linewidth = 0.5) +
  geom_tile(aes(x = (Begin + End)/2, y = Pho4, width = (End - Begin + 1)), 
            data = bHLH, height = h, fill = "yellowgreen") +
  scale_fill_manual("Activation", values = act.colors) +
  theme_cowplot() +
  theme(
    axis.line = element_blank(),
    axis.title = element_blank(),
    legend.position = "none"
  )
ggsave("../output/img/20240324-diverse-Pho4-PADDLE-for-R35.png", p3, width = 4, height = 3)  
```

The `echo: false` option disables the printing of code (only output is displayed).
