---
title: "Pho4-mNeon plasmids nuc vs wc or nuc vs cyto"
author: "JY, modified by HB"
date: "02/05/2024"
output: 
  html_notebook:
    toc: yes
    code_folding: hide
---

```{r message=FALSE}
library(tidyverse)
library(cowplot)
library(ggtext)
```

# Goal

The goal is to test whether the identity of the NLS region (region 3) affects the nuclear vs cytoplasmic ratio of the Pho4 constructs.

# Method
Mid-log phase _S.cerevisiae_ strains carrying Pho4 constructs with NLS:Sc or NLS:Cg were grown in SC complete medium. Live fluorescence microscopy was performed on a Leica epi-fluorescence microscope (Bright Field, GFP, DAPI). The TIFF image files were loaded into FIJI. Nuclei and cell boundaries were manually traced. Two quantities were calculated:

1. Fraction of nuclear Pho4 molecules.
    Integrated intensities in the nucleus vs whole cell were quantified. Non fluorescent cells were included in the second attempt to provide background correction (for autofluorescence). The ratio of nuclear vs cytoplasmic (= whole cell - nuclear) were used for comparisons.
2. Nuclear Pho4 concentration.
    The average gray intensity within the nucleus is quantified, and background was subtracted.

# Data and analysis
Two biological replicates with Pho2: d013024 and d020624; two biological replicates w/o Pho2: d021424 and d021624

The first replicate with Pho2 on d013024 didn't have a non-fluorescent cell control. In some of the analyses, we will use just one replicate.

File format:

- Pair_No: id for cell nested within "Group"
- Area: area in pixels
- Mean: mean fluorescence intensity, = RawIntDen / Area
- IntDen: Mean * area in scaled units, ignore for this analysis
- Median: median fluorescence intensity
- RawIntDen: sum of pixel values in the selected area
- Group: genotype
- Locus: area selected

## data import
```{r}
my_data_import <- function(file){
  # the goal is to calculate nuclear-vs-whole-cell ratio of the integrated GFP signal
  # also of interest is the nuclear concentration (mean or median of nucleus)
  # read the raw file
  raw <- read_tsv(file)
  # calculate the quantity of interest
  ratio <- raw %>% 
    # remove the cytoplasm rows, which can be calculated from the wc and nuclei
    filter(Locus != "cytoplasm") %>% 
    rename(Sum = RawIntDen) %>% 
    pivot_wider(id_cols = c(Group, Pair_No),
                names_from = Locus, values_from = c(Area, Sum)) %>% 
    mutate(Nuc_frac = Sum_nuclei / Sum_wc)
  
  median <- raw %>% 
    filter(Locus == "nuclei") %>% 
    select(Group, Pair_No, Median_nuc = Median)
  
  dat <- full_join(ratio, median, by = c("Group", "Pair_No"))
}
```

```{r message=FALSE}
samples <- read_tsv("../input/sample-list.txt", col_types = "cccc")
files <- dir("../input", pattern = "*.tsv")
tmp <- str_split(files, pattern = "-", simplify = TRUE)
files <- paste0("../input/", files)
names(files) <- paste(tmp[,1], tmp[,2], sep = "_")
dat <- map_dfr(files,
               \(f) my_data_import(f), .id = "file") %>% 
  separate(file, into = c("Date", "Host"), sep = "_") %>% 
  left_join(samples, by = c("Date", "Host", "Group")) %>% 
  mutate(
    Pho4 = fct_relevel(Pho4, "SSSSS", "CCSCC", "SSSCC",
                       "CCCCC", "SSCCS", "SSCSS"),
    NLS = substr(Pho4, 3, 3) %>% fct_relevel("S"),
    Host = fct_relevel(Host, "PHO2")
    #Rep = fct_recode(Date, A = "20240130", B = "20240206", A = "20240214", B = "20240216")
  )

# all sample names match?
all(!is.na(dat$Pho4))
```
Export the combined data
```{r}
write_tsv(dat, file = "../output/20250213-NLS-nuc-frac-data.tsv")
```

## Plot
```{r}
# labeller
par.name <- c(
  Nuc_frac = "Nuclear fraction",
  Median_nuc = "Nuclear concentration (a.u.)"
)

host.name <- c(
  PHO2 = "PHO2",
  #boost = "Boost",
  pho2 = "pho2∆"
)
my_symbol_colors <- function(symbols){
  x = str_sub(symbols, 3, 3) # NLS identity
  str_sub(symbols, 3, 3) <- ifelse(
    x == "C", 
    "<b style='color:blue'>C</b>", 
    "<b style='color:forestgreen'>S</b>")
  return(symbols)
}
p <- dat %>%
  filter(Date %in% c("20240206", "20240216")) %>% 
  select(Date, Host, Group, Nuc_frac, Median_nuc, Pho4, NLS) %>% 
  pivot_longer(cols = c(Nuc_frac, Median_nuc), 
               names_to = "Parameter", values_to = "Value") %>% 
  ggplot(aes(y = Pho4, x = Value)) +
  geom_boxplot(outliers = FALSE) +
  geom_point(color = "salmon",
             position = position_jitter(height = 0.1), 
             alpha = 0.8, size = 0.7) +
  geom_hline(yintercept = 3.5, linetype = 3, linewidth = 1.2, color = "gray40") +
  #facet_wrap(~Host, labeller = labeller(Host = host.name)) +
  facet_grid(Host ~ Parameter, scales = "free_x",
             labeller = labeller(Parameter = par.name, Host = host.name)) +
  scale_y_discrete(labels = my_symbol_colors, limits = rev) +
  theme_cowplot() + panel_border(color = "gray30") +
  background_grid(major = "x", minor = "none") +
  theme(axis.text.y = element_markdown(family = "courier"),
        axis.title = element_blank(),
        axis.line = element_blank(),
        strip.placement = "outside",
        strip.background = element_blank(),
        strip.text = element_text(face = 3, size = rel(1)))
p
ggsave("../output/20241012-NLS-identity-not-impact-nuc-frac.png", width = 6, height = 4)
```
## Statistical test
### all data, nuclear concentration (median_nuc)
ANOVA
```{r}
lm <- lm(Median_nuc ~ NLS + Host, data = dat)
anova(lm)
```

Non-parametric test
```{r}
print("Kruskal-Wallis test for PHO2 background")
kruskal.test(Median_nuc ~ NLS, data = dat, subset = dat$Host == "PHO2")
print("Kruskal-Wallis test for pho2∆ background")
kruskal.test(Median_nuc ~ NLS, data = dat, subset = dat$Host == "pho2")
```
### one day only
ANOVA
```{r}
tmp <- dat %>% filter(Date %in% c("20240206", "20240216"))
lm <- lm(Nuc_frac ~ NLS + Host, data = tmp)
anova(lm)
```
Non-parametric test
```{r}
print("Kruskal-Wallis test for PHO2 background")
kruskal.test(Nuc_frac ~ NLS, data = tmp, subset = dat$Host == "PHO2")
print("Kruskal-Wallis test for pho2∆ background")
kruskal.test(Nuc_frac ~ NLS, data = tmp, subset = dat$Host == "pho2")
```

