---
title: "Plot Y1H for activation potential test results"
author: Bin He
date: "2024-4-2 (updated `r Sys.Date()`)"
output:
  html_notebook:
    theme: cerulean
    toc: true
    toc_float: true
    toc_depth: 4
    code_folding: hide
---

```{r setup, message=FALSE}
require(tidyverse)
require(cowplot)
require(broom)
set.seed(12345)
old <- theme_set(theme_cowplot())
```

# Goal

Plotting Y1H data for testing the activation potential of PADDLE predicted regions in both ScPho4 and CgPho4

# Data

flow cytometry experiments were performed by Lindsey on 2024/01/11-15. gating was done on the Attune software and the stats were exported and organized

```{r message=FALSE}
sample <- read_tsv("../input/20240111-sample-list.txt", col_types = "cc", na = character()) %>% 
  mutate(name = fct_inorder(name), group = fct_inorder(group))
day1 <- read_csv("../input/20240111-Y1H-test-activation-potential-day1.csv") %>% 
  separate(sample, into = c("plasmid", "host", "rep")) %>% 
  left_join(sample, by = "plasmid") %>% 
  filter(!(plasmid == "NA" & host %in% c("blank", "549")))
day2 <- read_csv("../input/20240115-Y1H-test-activation-potential-day2.csv") %>% 
  separate(sample, into = c("plasmid", "host", "rep")) %>% 
  left_join(sample, by = "plasmid") %>% 
  filter(!(plasmid == "NA" & host %in% c("blank", "549")))
dat1 <- filter(day1, parameter == "GAL1pr-mCherry-H") %>% select(-parameter)
dat2 <- filter(day2, parameter == "GAL1pr-mCherry-H") %>% select(-parameter)
dat <- bind_rows(list("01/11" = dat1, "01/15" = dat2), .id = "date")
```

# Analysis

## Plot all data
```{r}
grp.colors <- c("control" = "gray50", 
                "ScPho4" = rgb(56, 118, 29, maxColorValue = 255), 
                "CgPho4" = rgb(40, 130, 200, maxColorValue = 255),
                "Combo" = rgb(142, 120, 200, maxColorValue = 255))
pd <- position_dodge(0.9)
p0 <- dat %>% 
  filter(name != "Gal4") %>% 
  ggplot(aes(x = name, y = median, group = date)) +
  geom_bar(aes(fill = group, alpha = date), stat = "summary", fun = "mean",
           position = pd) +
  stat_summary(fun.data = "mean_sdl", fun.args = list(mult = 1), 
               geom = "errorbar", width = 0.5, position = pd) +
  coord_flip() + 
  scale_x_discrete(limits = rev) +
  scale_alpha_manual(values = c(1, 0.5)) +
  scale_fill_manual(values = grp.colors)
p0
```

## Individual region tests
```{r}
base <- dat1 %>% filter(name %in% c("host", "Gal4DBD")) %>% pull(median) %>% mean()
tmp <- dat1 %>% filter(name != "Gal4", group != "Combo")
p1 <- tmp %>% 
  ggplot(aes(x = name, y = median)) +
  stat_summary(fun.data = "mean_sdl", fun.args = list(mult = 1), 
               geom = "errorbar", width = 0.5) +
  geom_bar(aes(fill = group), stat = "summary", fun = "mean") +
  geom_hline(yintercept = base, linetype = 2) +
  coord_flip() + 
  scale_x_discrete(limits = rev) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.15))) +
  scale_fill_manual(values = grp.colors, guide = "none") +
  panel_border(color = "gray20", size = 1.2) + 
  background_grid(major = "x", minor = "x") + 
  xlab(NULL) + ylab("GAL1pr-mCherry (a.u.)") +
  theme(
    axis.line = element_blank(),
    axis.text.y = element_text(size = rel(0.8))
  )
  
p1
ggsave("../output/20240402-single-region-Gal4DBD.png", width = 4, height = 3.8)
```
Statistical tests
```{r}
tmp1 <- mutate(tmp, name = fct_recode(name, bg = "host", bg = "Gal4_DBD")) %>% 
  droplevels()
lm <- lm(median ~ name, data = tmp1)
lm.res <- summary(lm)$coef
# remove the intercept term and standard error, which is the same for all levels
lm.res <- lm.res[-1,] %>% as.data.frame() %>% 
  rownames_to_column(var = "Region") %>% 
  mutate(
    Region = gsub("name", "", Region),
    P.adj = p.adjust(`Pr(>|t|)`, method = "holm"),
    Sig = cut(P.adj, breaks = c(0, 0.001, 0.01, 0.05, 0.1, 1),
              labels = c("***", "**", "*", ".", " "))
  )
print(lm.res[, -3], digits = 3)
```

## Booster effect
```{r}
#base <- dat1 %>% filter(name %in% c("host", "Gal4DBD")) %>% pull(median) %>% mean()
select <- c("host", "Gal4_DBD", "ScAD", "ScE1:ScAD", "CgAD", "CgE1:CgAD", 
            "ScAD_9aa", "CgE1:ScAD_9aa", "ScAD_9aa:CgE1",
            "CgE2_9aa:ScAD_9aa", "ScAD_9aa:CgE2_9aa"
            )
tmp <- dat1 %>% filter(name %in% select) %>% droplevels() %>% 
  mutate(name = factor(name, levels = select))
p2 <- tmp %>% 
  ggplot(aes(x = name, y = median)) +
  stat_summary(fun.data = "mean_sdl", fun.args = list(mult = 1), 
               geom = "errorbar", width = 0.5) +
  geom_bar(aes(fill = group), stat = "summary", fun = "mean", width = 0.8) +
  #geom_hline(yintercept = base, linetype = 2) +
  geom_vline(xintercept = c(5.5, 7.5, 9.5),
             linetype = 3, linewidth = 0.7, color = "gray20") +
  coord_flip() + 
  scale_x_discrete(limits = rev) +
  scale_y_continuous(breaks = c(0, 2000, 4000),
                     expand = expansion(mult = c(0.02, 0.15))) +
  scale_fill_manual(values = grp.colors, guide = "none") +
  panel_border(color = "gray20", size = 1.2) + 
  background_grid(major = "x", minor = "x") + 
  xlab(NULL) + ylab("GAL1pr-mCherry (a.u.)") +
  theme(
    axis.line = element_blank(),
    axis.text.y = element_text(size = rel(0.8))
  )
  
#p2
p <- plot_grid(p1, NULL, p2, nrow = 1, rel_widths = c(1, 0.1, 1.2))
p
ggsave("../output/20240402-booster-region-Gal4DBD.png", width = 7, height = 3.6)

```

Statistical test
```{r}
# ScE1 on ScAD
x <- dat1$median[dat1$name == "ScE1:ScAD"]
y <- dat1$median[dat1$name == "ScAD"]
test1 <- t.test(x, y) 

# CgE1 on CgAD
x <- dat1$median[dat1$name == "CgE1:CgAD"]
y <- dat1$median[dat1$name == "CgAD"]
test2 <- t.test(x, y)

# Different combination on ScAD_9aa
subset <- select[7:length(select)]
tmp1 <- dat1 %>% filter(name %in% subset) %>% droplevels() %>% 
  mutate(name = factor(name, levels = subset))
lm <- lm(median ~ name, data = tmp1)
lm.res <- summary(lm)$coef[-1,] # remove intercept term (=ScAD_9aa mean)
# extract the name and p-values (raw)
p.values <- tibble(
  region = rownames(lm.res),
  rawP = lm.res[, "Pr(>|t|)"]
) %>% 
  add_row(region = "ScE1:ScAD", rawP = test1$p.value) %>% 
  add_row(region = "CgE1:CgAD", rawP = test2$p.value) %>% 
  mutate(
    P.adj = p.adjust(rawP, method = "holm"),
    Sig = cut(P.adj, breaks = c(0, 0.001, 0.01, 0.05, 0.1, 1),
              labels = c("***", "**", "*", ".", " "))
  )

p.values
```


Use contrast and coding schemes (experimental)
```{r}
contr_mat <- matrix(c(0, 0, -1, 1, rep(0, 7),
                      0, 0, 0, 0, -1, 1, rep(0, 5),
                      rep(0, 6), -1, 1, 0, 0, 0,
                      rep(0, 6), -1, 0, 1, 0, 0,
                      rep(0, 6), -1, 0, 0, 1, 0,
                      rep(0, 6), -1, 0, 0, 0, 1),
                    nrow = 6, byrow = TRUE)
rownames(contr_mat) <- p.values$region

lm <- lm(median ~ name, data = tmp, 
         contrasts = list(name = MASS::ginv(contr_mat)))
summary(lm)
```

