---
title: "PHO5 mutant promoter reporter assay"
date: 2025-04-26, updated `r Sys.Date()`
author: "Bin He"
format: html
---

```{r}
require(tidyverse)
require(cowplot)
require(ggtext)
```

```{r}
old <- theme_set(theme_cowplot())
```

## Goal

Plot flow cytometry result for _PHO5_ wild type (WT) and mutant reporter data.

## Data

The "quick" analysis uses data pre-processed by Kelsey. She exported the MFI from the Attune software, then did background subtraction for both Pho4-mNeon and _PHO5pr_-mCherry, then calculated the ratio between the two, which was used to describe the activity of the TF.

```{r}
dat <- read_tsv("../input/20250425-PHO5-mutant-reporter-KM-excel-export.tsv", 
                na = "NA") |>
  pivot_longer(cols = X1:X6, names_to = "Rep", values_to = "Activity") |>
  mutate(PHO5pr = factor(PHO5pr, c("WT", "M1", "M2")))
```

## Plot

```{r}
p <- dat %>% 
  filter(Host == "PHO2") %>%
  ggplot(aes(x = PHO5pr, y = Activity)) +
  geom_bar(stat = "summary", fun = "mean", 
           fill = "gray80", color = "gray20") +
  geom_point(shape = 3) +
  facet_wrap(~Pho4) +
  xlab("_PHO5_ promoter variant") + ylab("Normalized TF activity") +
  theme(strip.background = element_blank(),
        strip.text = element_text(size = rel(1)),
        axis.title = element_markdown())
p
ggsave("../output/20250426-PHO5-mutant-promoter-induction-both-Pho4.png",
       width = 5, height = 2.5)
```

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
